---
title: 'Edward Example: Dirichlet Process'
output:
  html_document:
    toc: yes
  html_notebook:
    toc: yes
---

## Preface

This is an R implementation of the Edward example [Python script for Dirichlet process priors](https://github.com/blei-lab/edward/blob/master/examples/pp_dirichlet_process.py).

(*Implemented under Edward's [license instructions](http://edwardlib.org/license), i.e., [Apache License, version 2.0](https://opensource.org/licenses/Apache-2.0).*)

## Setup

First, load the R packages and then import the Edward module into the underlying Python system.

```{r package, message=FALSE, warning=FALSE}

library( magrittr )
library( tidyverse )
library( reticulate ) # R interface to Python
library( tensorflow )

ed   <- import( module = "edward" )

#from edward.models import 
Bernoulli         <- ed$models$Bernoulli
Beta              <- ed$models$Beta
DirichletProcess  <- ed$models$DirichletProcess
Exponential       <- ed$models$Exponential
Normal            <- ed$models$Normal
MvNormalDiag      <- ed$models$MultivariateNormalDiag
```


## Dirichlet process.

We implement sample generation from a Dirichlet process (with no base
distribution) via its stick breaking construction. It is a streamlined
implementation of the ``DirichletProcess`` random variable in Edward.

### References

[Probabilistic Models of Cognition: Chapter 12, Non-parametric Models](https://probmods.org/chapters/12-non-parametric-models.html#infinite-discrete-distributions-the-dirichlet-processes) by Noah D. Goodman & Joshua B. Tenenbaum 


## Stick-Breaking Construction

First, demonstrate Stick-breaking process as a function.

```{r def_funcs}

dirichlet_process <- function( alpha ){
  #"""Demo of stochastic while loop for stick breaking construction."""
  cond <- function( k, beta_k ){
    # End while loop (return False) when flip is heads.
    flip <- Bernoulli( beta_k )
    return( tf$logical_not( tf$cast( flip , tf$bool ) ) )
  }

  body <- function( k, beta_k ){
    beta_k <- Beta( 1.0, alpha )
    return( list( k + 1L, beta_k ) )
  }

  k      <- tf$constant( 0L )
  beta_k <- Beta( 1.0, alpha )
  result <- tf$while_loop( cond, body, loop_vars = list( k, beta_k ) )
  return( result )
}
```

```{r make_dp}
tf$reset_default_graph()
ed$set_seed( 42L )

dp <- dirichlet_process( alpha = 10.0 )

# The number of sticks broken is dynamic, changing across evaluations.
#sess <- tf$Session()
sess <- ed$get_session()
print( sess$run( dp ) )
print( sess$run( dp ) )
```

## Edward's `DirichletProcess` Random Variables

Now, demonstrate the built-in Edward `DirichletProcess` random variables.


```{r demo_dp}
# Demo of the DirichletProcess random variable in Edward.
base <- Normal(0.0, 1.0)

N_sample <- 10000L

# Highly concentrated DP.
alpha <- 1.0

dp    <- DirichletProcess(alpha, base)
x     <- dp$sample( N_sample )
samples <- sess$run(x)

data_frame( samples = samples ) %>%
{
ggplot( data = ., aes(x = samples) ) +
    geom_histogram( bins=100 ) +
    lims( x = c(-3.0, 3.0) ) +
    ggtitle( sprintf("DP(%g, N(0, 1))", alpha) )
} %>% 
  print()
```

```{r spread_dp}
# More spread out DP.
alpha   <- 50.0

dp      <- DirichletProcess(alpha, base)
x       <- dp$sample( N_sample )
samples <- sess$run(x)
data_frame( samples = samples ) %>%
{
ggplot( data = ., aes(x = samples) ) +
    geom_histogram( bins=100 ) +
    lims( x = c(-3.0, 3.0) ) +
    ggtitle( sprintf("DP(%g, N(0, 1))", alpha) )
} %>% 
  print()
```


States persist across calls to sample() in a DP.

```{r persistence}
# States persist across calls to sample() in a DP.
alpha <- 1.0
dp <- DirichletProcess(alpha, base)
N_samp <- c(samples_x = 50L, samples_y = 75L )
x  <- dp$sample(N_samp[1])
y  <- dp$sample(N_samp[2])
result <- sess$run(list(x, y)) %>% setNames( names(N_samp))

data_frame( 
  samples = do.call(c,result), 
  label = rep(names(N_samp),times=N_samp) 
) %>%
{
ggplot( data = ., aes(x = samples ) ) +
    geom_histogram( bins = 100 ) +
    lims( x = c(-3.0, 3.0) ) +
    ggtitle( sprintf("DP(%g, N(0, 1))", alpha) ) +
    facet_wrap( ~label)
} %>% 
  print()

```



The random variable `theta` is the distribution indirectly returned by the DP.
Fetching `theta` is the same as fetching the Dirichlet process.

```{r induced }
# ``theta`` is the distribution indirectly returned by the DP.
# Fetching theta is the same as fetching the Dirichlet process.
dp <- DirichletProcess(alpha, base)
theta <- Normal(0.0, 1.0, value=tf$cast(dp, tf$float32))
print(sess$run(list(dp, theta)))
print(sess$run(list(dp, theta)))
```

DirichletProcess can also take in non-scalar concentrations and bases.

```{r nonscalar }
# DirichletProcess can also take in non-scalar concentrations and bases.
alpha <- tf$constant(list(0.1, 0.6, 0.4))
base <- Exponential( rate = tf$ones( c(5L, 2L) ) )
dp <-  DirichletProcess(alpha, base)
print(dp)
print(sess$run(dp))
```


## Multivariate Base

{MLT}: Does it work with a multivariate base distribution?

```{r multivar}

# Demo of the DirichletProcess random variable in Edward.
base <- MvNormalDiag(
  loc = c(1., -1),
  scale_diag = c(1, 2.)
)

N_sample <- 10000L

alpha <- 1.0/100.0

dp    <- DirichletProcess( alpha, base )
x     <- dp$sample( N_sample )
samples <- sess$run(x)

samples %>%
  as_data_frame() %>%
  setNames( c( "X1","X2" ) ) %>%
{
ggplot( data = ., aes(x = X1, y = X2 ) ) +
    stat_density2d( aes(alpha=..level..), geom="polygon") +
    geom_point( color = 'red', alpha = 0.3 ) +
    lims( x = c(-5.0, 5.0), y = c(-5.0, 5.0 ) ) +
    ggtitle( sprintf("DP(%g, MvN({1,-1}, diag({1,2})))", alpha) )
} %>% 
  print()

```



