---
title: 'R Markdown port of "Edward: Linear Mixed Effects Models" tutorial'
author: "Michael L. Thompson"
date: "July 30, 2017"
output:
  pdf_document:
    toc: yes
  html_notebook:
    toc: yes
  html_document:
    toc: yes
linkcolor: red
urlcolor: blue
---

## Preface

This is an R implementation of a variation of the Edward tutorial ["Getting Started"](http://nbviewer.jupyter.org/github/blei-lab/edward/blob/master/notebooks/getting_started.ipynb) (jupyter notebook).  

*Done per Edward's [license instructions](http://edwardlib.org/license), i.e., [Apache License, version 2.0](https://opensource.org/licenses/Apache-2.0).*


# Linear Mixed Effects Models

With linear mixed effects models, we wish to model a linear
relationship for data points with inputs of varying type, categorized
into subgroups, and associated to a real-valued output.

We demonstrate with an example in Edward. A webpage version is available 
[here](http://edwardlib.org/tutorials/linear-mixed-effects-models).

```{r packages, message=FALSE, warning=FALSE}
library( magrittr )
library( tidyverse )
library( reticulate )
library( tensorflow )

ed <- import( module = "edward" )
#pd <- import( module = "pandas" )
np <- import( module = "numpy" )

# Assign the Edward models used in the script.
Normal <- ed$models$Normal

ed$set_seed(42L)

```

## Data

We use the `InstEval` data set from the popular
[lme4 R package](http://lme4.r-forge.r-project.org) (Bates, Mächler, Bolker, & Walker, 2015).
It is a data set of instructor evaluation ratings, where the inputs
(covariates) include categories such as `students` and
`departments`, and our response variable of interest is the instructor
evaluation rating.

```{r data}
# s - students - 1:2972
# d - instructors - codes that need to be remapped
# dept also needs to be remapped
data <- read_csv('data/insteval.csv')
data %<>% 
  rename( index = X1 ) %>%
  mutate(
    dcodes    = factor(sprintf("%04d",d)),
    deptcodes = factor(sprintf("%02d",dept)),
    s         = s - 1
  ) %>%
  select( index, everything() )

train <- data %>% sample_frac( 0.8, replace = FALSE )
test  <- data %>% filter( !(index %in% train$index) )

train %>% print()

```
In the code, we denote:
+ `students` as `s`
+ `instructors` as `d`
+ `departments` as `dept`
+ `service` as `service`

```{r preproc}
sub1L <- function(x) as.integer(x) - 1L
train_new <- train  %>% 
  select(index,s,dcodes,deptcodes,service,y) %>% 
  mutate_at( .funs=funs( sub1L ) , vars(-index,-service,-y) ) %>%
  mutate( y = as.numeric( y ) )

n_obs_train <- nrow( train_new )

test_new <- test  %>% 
  select(index,s,dcodes,deptcodes,service,y) %>% 
  mutate_at( .funs=funs( sub1L ) , vars(-index,-service,-y) ) %>%
  mutate( y = as.numeric( y ) )

n_obs_test <- nrow( test_new )

```

```{r counts}
n_s    <- 2972L        # n_s    = length(unique(data$s)) # number of students
n_d    <- 1128L        # n_d    = length(unique(data$d)) # number of instructors
n_dept <- 14L          # n_dept = length(unique(data$dept)) # number of departments
n_obs  <- nrow(train) # number of observations

```
## Model

With linear regression, one makes an independence assumption where
each data point regresses with a constant slope among
each other. In our setting, the observations come from
groups which may have varying slopes and intercepts. Thus we'd like to
build a model that can capture this behavior (Gelman & Hill, 2006).

For examples of this phenomena:
+ The observations from a single student are not independent of
each other. Rather, some students may systematically give low (or
high) lecture ratings.
+ The observations from a single teacher are not independent of
each other. We expect good teachers to get generally good ratings and
bad teachers to get generally bad ratings.
+ The observations from a single department are not independent of
each other. One department may generally have dry material and thus be
rated lower than others.


Typical linear regression takes the form

\begin{equation*}
\mathbf{y} = \mathbf{X}\beta + \epsilon,
\end{equation*}

where $\mathbf{X}$ corresponds to fixed effects with coefficients
$\beta$ and $\epsilon$ corresponds to random noise,
$\epsilon\sim\mathcal{N}(\mathbf{0}, \mathbf{I})$.

In a linear mixed effects model, we add an additional term
$\mathbf{Z}\eta$, where $\mathbf{Z}$ corresponds to random effects
with coefficients $\eta$. The model takes the form

\begin{align*}
\eta &\sim \mathcal{N}(\mathbf{0}, \sigma^2 \mathbf{I}), \\
\mathbf{y} &= \mathbf{X}\beta + \mathbf{Z}\eta + \epsilon.
\end{align*}

Given data, the goal is to infer $\beta$, $\eta$, and $\sigma^2$,
where $\beta$ are model parameters ("fixed effects"), $\eta$ are
latent variables ("random effects"), and $\sigma^2$ is a variance
component parameter.

Because the random effects have mean 0, the data's mean is captured by
$\mathbf{X}\beta$. The random effects component $\mathbf{Z}\eta$
captures variations in the data (e.g.  Instructor \#54 is rated 1.4
points higher than the mean).

A natural question is the difference between fixed and random effects.
A fixed effect is an effect that is constant for a given population. A
random effect is an effect that varies for a given population (i.e.,
it may be constant within subpopulations but varies within the overall
population). We illustrate below in our example:

+ Select `service` as the fixed effect. It is a binary covariate
corresponding to whether the lecture belongs to the lecturer's main
department. No matter how much additional data we collect, it
can only take on the values in $0$ and $1$.
+ Select the categorical values of `students`, `teachers`,
and `departments` as the random effects. Given more
observations from the population of instructor evaluation ratings, we
may be looking at new students, teachers, or departments.

In the syntax of R's lme4 package (Bates et al., 2015), the model
can be summarized as

```
y ~ 1 + (1|students) + (1|instructor) + (1|dept) + service
```
where `1` denotes an intercept term,`(1|x)` denotes a
random effect for `x`, and `x` denotes a fixed effect.

```{r}
# Set up placeholders for the data inputs.
s_ph       <- tf$placeholder(tf$int32, NULL)
d_ph       <- tf$placeholder(tf$int32, NULL)
dept_ph    <- tf$placeholder(tf$int32, NULL)
service_ph <- tf$placeholder(tf$float32, NULL)

# Set up fixed effects.
mu      <- tf$Variable(tf$random_normal(shape()))
service <- tf$Variable(tf$random_normal(shape()))

sigma_s    <- tf$sqrt(tf$exp(tf$Variable(tf$random_normal(shape()))))
sigma_d    <- tf$sqrt(tf$exp(tf$Variable(tf$random_normal(shape()))))
sigma_dept <- tf$sqrt(tf$exp(tf$Variable(tf$random_normal(shape()))))

# Set up random effects.
eta_s <- Normal(loc=tf$zeros(n_s), scale=sigma_s * tf$ones(n_s))
eta_d <- Normal(loc=tf$zeros(n_d), scale=sigma_d * tf$ones(n_d))
eta_dept <- Normal(loc=tf$zeros(n_dept), scale=sigma_dept * tf$ones(n_dept))

# (gather(tensor,indices) selects the values of the specified indices from the tensor)
yhat <-
  tf$gather(eta_s, s_ph) + tf$gather(eta_d, d_ph) + tf$gather(eta_dept, dept_ph) + mu + service*service_ph

y <- Normal(loc=yhat, scale=tf$ones(n_obs))
```
